---
title: "RSQLite Example Queries"
author: "Ian Eykamp"
modified by" Rodrigo Gonzalez-Abraham
date: "11/29/2021"
mod date: "06/30/2025/
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## 4.3.1 Setup

Run the code chunk below to connect to the `latest.db` database file.
We will run queries on the Sensorbot database using SQL commands.

```{r}
# load the data from the Sensorbot database
load("C:/dev/sensorbot/sensorbot.RData")

# install or load required packages using pacman
if (!require("pacman")) install.packages("pacman")
pacman::p_load(pacman, RSQLite, knitr)

conn <- dbConnect(
  drv = RSQLite::SQLite(), 
  dbname = "C:/dev/sensorbot/sensorbot.db"
) 
```

Note, SQL code is different from R code: users cannot type SQL code
directly into RStudio. We will use an R function called `dbGetQuery` to
perform SQL queries on the database. `dbGetQuery` takes as arguments a
connection to the database (defined above) and a string containing the
SQL syntax we wish to execute, like so:

SQL syntax is not sensitive to whitespace, so we can add as many spaces
or line breaks to our string as we want. Just remember, don't put
punctuation or commas unless it is part of the SQL syntax.

```{r}
dbGetQuery(conn, "SELECT * FROM device LIMIT 10")
```

## 4.3.2 Make Table Command

There are at least two ways to create a new table in SQL. You can
initialize one from scratch by specifying the names of each of the
columns you want to create. Each column is associated with a datatype
and an optional constraint, which dictates what kind of data can be
inserted into the table. For example, you might want to dictate up front
that you do not want duplicate values for a column (such as a database
storing different people's usernames); if this is the case, you might
want to use the `UNIQUE` constraint on the username column.

Another way to make a table is to copy data from one or more existing
tables. This is useful if you want to make a backup copy while you mess
around with the original data, or if you want to collect useful
information from across tables into one convenient place.

Note the following definitions for the Make a Table command:

-   `new_table_name`: Name of the table to be created.
-   `column1`: Name of the first column and so on.
-   `data_type`: Type of data that can be stored in the field.
-   `size`: Size of the data that can be stored in a column. For
    example, in a column, if you give the data_type as `varchar` and
    size as 20, that column can store a string of 20 characters,
    maximum.
-   `constraint_name`: Name of the constraint. For example, `NOT NULL`,
    `UNIQUE`, `PRIMARY KEY`.

One last thing to note is that we are using the `dbExecute` command (not
`dbGetQuery`), because we are making a change to the database (not
merely querying existing data. `dbExecute` returns either a 0 if the
command failed or a 1 if it succeeded.

Template:

```{r eval=FALSE}
# Template 1: creating a table from scratch by defining column names and types
dbExecute(conn, "

CREATE TABLE new_table_name ( 
  column1 data_type(size) constraint_name,
  column2 data_type(size) constraint_name, 
  column3 data_type(size) constraint_name, 
  .... 
)

")

##--------------------------------------------------------------------------------##

# Template 2: creating a table by copying another table or a subset of another table
dbExecute(conn, "

CREATE TABLE IF NOT EXISTS new_table_name
  AS 
SELECT column1, column2, column3, ... FROM existing_table_name

")
```

Example:

```{r}
# Example 1: creating a table from scratch by defining column names and types
dbExecute(conn, " 

CREATE TABLE IF NOT EXISTS ts_kv_example (
  entity_id STRING NOT NULL, 
  key INT NOT NULL, 
  ts INT NOT NULL, 
  val FLOAT NOT NULL
) 

")

##-------------------------------------------------------------------------------##

# Example 2: creating a table by copying another table or a subset of another table
dbExecute(conn, "

CREATE TABLE IF NOT EXISTS dictionary_example 
  AS 
SELECT * FROM ts_kv_dictionary

")
```

## 4.3.3 Insert Data Command

If you want to add a row of data to a table, use the `INSERT` command.
You need to provide one value for each column in the table, and the data
types must match the data types of the columns.

-   `table_name`: The name of the table.

-   `value1`, `value2`, `…`: The value of the first column, second
    column, and so on, for the new record.

Template:

```{r eval=FALSE}
dbExecute(conn, "

INSERT INTO table_name
  VALUES (value1, value2, value3,...)

")
```

Example:

```{r}
dbExecute(conn, "

INSERT INTO dictionary_example 
  VALUES ('rain', '60')

")
```

This added a row for "rain" into our dictionary, with a key
identification (`key_id`) of 60. Now try this function again, but add a
row for "wildfire" and label that `key_id` as 61. Then run the following
command.

```{r}
dbGetQuery(conn, "

SELECT * FROM dictionary_example
WHERE key_id = 60 OR key_id = 61

")
```

## 4.3.4 Select Statement

The `SELECT` statement is the most important command in SQL. It allows
you to formulate a query to retrieve anything from a single precise
piece of information to an entire table (or multiple tables). Any query
starts with `SELECT ... FROM`, and there are many options for refining
the query from there.

Use the following template if you want to select one or more specific
columns from a table. If you wish to select all columns from a table,
use `*` to represent "everything"

Template:

```{r eval=FALSE}
dbGetQuery(conn, "

SELECT column1, column2 FROM table_name

")
```

Example:

```{r}
dbGetQuery(conn, "

SELECT * FROM ts_kv_dictionary

")
```

## 4.3.5 Where Clause

You can use the `WHERE` clause to filter which rows to return from a
table. You can construct highly complex queries using logical operators
like `AND` and `OR`.

-   `column1` , `column2`: Fields in the table.

-   `tableName`: Name of the table.

-   `column_name`: Name of the field.

-   `operator`: Operation to be performed for filtering.

-   `value`: The exact value to get related data in a result.

Table 4-1 defines the operators that apply to the Where Clause Command.

| Operator |                Description                 |
|:--------:|:------------------------------------------:|
|    \>    |                greater than                |
|   \>=    |          greater than or equal to          |
|    \<    |                 less than                  |
|   \<=    |           less than or equal to            |
|    =     |                  equal to                  |
|   \<\>   |                not equal to                |
| BETWEEN  |           in an inclusive range            |
|   LIKE   |            search for a pattern            |
|    IN    | define multiple likely values for a column |

Template:

```{r eval=FALSE}
dbGetQuery(conn, "

SELECT column1,column2 
FROM TableName 
WHERE column_name [OPERATOR] value

")
```

Example:

```{r}
# note that "--" denotes a comment in SQL
dbGetQuery(conn, "

SELECT * FROM ts_kv_hourly
WHERE key = 10
AND ts > 1577880000000 -- timestamps after Jan 1, 2020

")
```

## 4.3.6 Limit

The previous query returned upwards of 250,000 rows. If we don't need
all this data, we can set a `LIMIT` on how many rows are returned. A
limit is often useful when you are just exploring or are in the process
of refining a complex query on a large database.

A note of caution: SQL returns rows in arbitrary order, so the 100 rows
that we get from this example are not necessarily the "first" 100 rows
in the database. To get top 100 values ranked in a certain order, use
`ORDER BY` in the next section.

Template:

```{r eval=FALSE}
dbGetQuery(conn, "

SELECT column1,column2 
FROM table_name
WHERE 
  ...

LIMIT number

")
```

Example:

```{r}
dbGetQuery(conn, "

SELECT * FROM ts_kv_hourly
WHERE key = 10
AND ts > 1577880000000 -- timestamps after Jan 1, 2020
LIMIT 100

")
```

## 4.3.7 Order By

Use the `ORDER BY` clause to sort the results of a query based on the
values in a specific column. For example, if we want to arrange the
entries according to the timestamp, we would use the following query.
Note that the `ORDER BY` clause comes before the `LIMIT`, if you are
using one.

You can sort in ascending or descending order (this determines whether
the highest or lowest value appears first). The default method is
ascending (low to high). If you wish to sort in descending order (high
to low), then use the keyword `DESC` after the name of the column you
are sorting by.

Template:

```{r eval=FALSE}
dbGetQuery(conn, "

SELECT column1,column2 
FROM table_name
WHERE 
  ...

ORDER BY column1
LIMIT number

")
```

Example:

```{r}
dbGetQuery(conn, "

SELECT * FROM ts_kv_hourly
WHERE key = 10
AND ts > 1577880000000 -- timestamps after Jan 1, 2020
ORDER BY ts DESC
LIMIT 100

")
```

## 4.3.8 Min/Max/Average/Total Number

The `MIN`, `MAX`, `AVG`, and `COUNT` commands return a single value ---
the minimum, maximum, average, and of the elements in a query,
respectively. You can be as specific as you like with the FROM and WHERE
clauses, and you will receive only the `MIN`, `MAX`, `AVG`, or `COUNT`
of those values you selected.

```{r eval=FALSE}
dbGetQuery(conn, "

SELECT MAX( column_name )
FROM table_name
WHERE ...

")
```

Example:

```{r}
dbGetQuery(conn, "

SELECT MAX(val) 
FROM ts_kv_hourly 
WHERE key = 7

")
```

## 4.3.9 Distinct Entries

A similar query is to find the distinct, or unique, values in a query.
Suppose you want to know which sensors are present in the `ts_kv_hourly`
table. For this, use the `DISTINCT` command in the same location as you
would `MIN`, `MAX`, `COUNT`, or `AVG`.

Template:

```{r eval=FALSE}
dbGetQuery(conn, "

SELECT DISTINCT( column_name )
FROM table_name 
WHERE ...
 
")
```

Example:

```{r}
dbGetQuery(conn, "

SELECT DISTINCT( entity_id )
FROM ts_kv_hourly

")
```

## 4.3.10 Combination Queries

In this section, we will cover five ways to use combination queries to
pull data that match search criteria based on genuine questions about
the database. This will combine the queries we covered in previous
sections.

#### 1. Count the number of readings where the particulate matter (PM)2.5 is greater than 500.

```{r}
dbGetQuery(conn, "

SELECT COUNT(*) AS total 
FROM ts_kv_hourly 
WHERE key = 10 AND val > 500

")
```

#### 2. Select the PM2.5 and the temperature for September 10 to 30, 2020.

```{r eval=FALSE}
dbGetQuery(conn, "

SELECT * FROM ts_kv_hourly 
WHERE key = 13 OR 16 AND 
ts BETWEEN 1599696000000 AND 1601424000000

")
```

Since the timestamp format is given in epoch milliseconds, for the
BETWEEN values, use this website ([Epoch
time](https://www.epochconverter.com/)) to convert the time range to
epoch timestamps and copy the Timestamp in milliseconds result to your
query.

![Screenshot from
[epochconverter.com](https://www.epochconverter.com/)](epoch_time_converter.png)

For example, if you want to see only the data from between September 5
and 10, 2020, your last line would be:
`ts BETWEEN 159928200000 AND 1599721200000`;

#### 3. Find the identification codes for each device managed by the

    Department of Environmental Quality (DEQ).

    Hint: They will be designated under their own table

```{r}
dbGetQuery(conn, "

SELECT DISTINCT entity_id 
FROM ts_kv_deq

")
```

#### 4. Select the data from days where the temperature exceeded 100 degrees

    Fahrenheit (F).

```{r}
dbGetQuery(conn, "

SELECT * FROM ts_kv_hourly
WHERE key = 16 AND val > (100-32) * 5/9

")
```

#### 5. Select maximum and minimum readings for PM10 for August 31, 2021.

```{r}
dbGetQuery(conn, "

SELECT MIN(val), MAX(val) FROM ts_kv_hourly
WHERE key = 7 AND 
ts BETWEEN 1630368000000 AND 1630454400000

")
```
