---
title: 'Module 5: Correlation'
author: "Ian Eykamp"
Modified: "Rodrigo Gonzalez-A"
date: "10/12/2021"
Mod date: 06/30/2025
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r library}

# install or load required packages using pacman
if (!require("pacman")) install.packages("pacman")
pacman::p_load(pacman, tidyverse, magrittr, knitr)

knitr::opts_chunk$set(echo = TRUE)
```

Clone the repo from <https://github.com/rodrigoganoso/Blueprint.git>
and load the file "pmComparison.RData"

```{r load}
# (change the path below to the file you cloned.)
load("./pmComparison.RData")
```

The above chunk loads the dataframe pmComparison into memory.
pmComparison has a timestamp, a PM2.5 reading, and a sensor field. The
sensor field tells you whether the reading was made by a DEQ reference
monitor or a PMS5003 sensor.

```{r view}
# Take a look at the dataframe.
head(pmComparison)
print(size_sum(pmComparison))
print(paste(min(pmComparison$date), "to", max(pmComparison$date)))
```

Now create a plot of DEQ and PMS5003 data. The way the dataframe is
structured makes it easy to color the data points differently based on
the value of the sensor column.

```{r plot}
ggplot(pmComparison, aes(x = date, y = pm25, color = sensor)) + 
    geom_point() + 
    labs(x = "Date (2020)", y = "PM2.5 Concentration", color = "Sensor type")
```

To get ready for the next step, you must convert the dataframe into a
"wide" format. That is, create separate columns for the DEQ and PMS5003
sensor readings, so they can easily be plotted against each other. It is
a subtle difference, but understanding this will make a huge difference
in your ability to handle data.

```{r pivot}
pmCompareCols <- 
  pivot_wider(
    pmComparison, 
    names_from = sensor, 
    values_from = pm25
  )

# Take a look at the new dataframe.
head(pmCompareCols)
```

```{r compare}
# and continue your comparison code below...

```

*Hint: to create a new code chunk, press* `Ctrl+Alt+I`.
